---
title: "Individual Assignment 5"
author: "Kelsey Hammond"
date: "05-10-2026"
format: pdf
execute: 
  message: false
  warning: false
---

# 1. Set up

## Packages

```{r}
#| label: packages

# general use and cleaning
library(tidyverse)
library(here)
library(janitor)
library(snakecase)

# wrapping axis labels
library(scales)

# reading in .xlsx files
library(readxl)

# visualizing missing data
library(naniar)

# arranging plots
library(patchwork)
```

## Data

```{r}
#| label: data

# taxonomic information
taxon_list <- read_csv(here("data", "taxon_list.csv")) 

# aquatic invertebrates
aq_ins <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                    sheet = "Aquatic Insects") 

# site information
sites <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                   sheet = "sites")

# water quality information
water_quality <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                           sheet = "Water Quality",
                           na = c("", "n/a", "N/A", "over", "173+")) 
```

## Cleaning Data

### New `sites` object

```{r}
#| label: ncos-site-object

# creating new object from sites
ncos_sites <- sites |> 
  # filter to only include NCOS sites sampled four times a year
  filter(sector == "NCOS" & sampling_frequency == "Quarterly")
```

### Cleaning `taxon_list`

```{r}
#| label: cleaning-taxon-list

# creating new clean object from taxon_list
taxon_list_clean <- taxon_list |> 
  # converting taxon name to snake case (no spaces or capital letters,
  # only underscores)
  mutate(verbatim_name = to_any_case(verbatim_name, case = "snake"))
```

### Cleaning `water_quality`

```{r}
#| label: water-quality-cleaning

# creating new clean object from water quality
water_quality_clean <- water_quality |> 
  # clean column names
  clean_names() |> 
  # filtering join: only include sites in ncos_sites
  semi_join(ncos_sites, by = "site") |> 
  # create new column to join with later data frames
  unite("date_site", date, site, remove = TRUE) |> 
  # group by date_site column
  group_by(date_site) |> 
  # calculate median pH, salinity, DO
  summarize(med_ph = median(p_h, na.rm = TRUE),
            med_sal = median(salinity_ppt, na.rm = TRUE),
            med_do = median(dissolved_oxygen_mg_l, na.rm = TRUE)) |> 
  # ungroup data frame
  ungroup() |> 
  # filter out salinity outliers
  filter(date_site != "2022-08-12_NVBR") 
```

### Cleaning `aq_ins`

```{r}
#| label: cleaning-aq-ins

# creating new clean object from aquatic inverts
aq_ins_clean <- aq_ins |> 
  # clean column names
  clean_names() |> 
  # filtering join: only include sites occurring in ncos_sites
  semi_join(ncos_sites, by = c("site")) |> 
  # renaming columns
  rename(date = date_on_vial) |> 
  # select columns of interest
  select(date, sample_type, site,
         ostracod,
         copepod,
         hemiptera_corixidae_boatman,
         cladocera,
         nematode,
         diptera_ceratopogonidae, 
         annelida_oligochaete,
         diptera_chironomid,
         annelida_polychaete) |> 
  # filter to only include "filtered beaker" samples
  filter(sample_type %in% c("FB 250", "FB250")) |> 
  # convert the data frame to long format
  pivot_longer(cols = ostracod:annelida_polychaete,
               names_to = "taxon",
               values_to = "abundance") |> 
  # group by date, site, taxon
  group_by(date, site, taxon) |> 
  # calculate average abundance per liter (sum abundance divided by 7.5 liters)
  summarize(ave_lit = sum(abundance, na.rm = TRUE)/7.5) |> 
  # ungroup data frame
  ungroup() |> 
  # create a new column called `date_site` to join with water quality
  unite("date_site", date, site, remove = FALSE) |> 
  # join with water quality
  left_join(water_quality_clean, by = c("date_site")) |> 
  # join with taxon list
  left_join(taxon_list_clean, by = c("taxon" = "verbatim_name")) |> 
  # add full names of sites
  mutate(site_full = case_when(
    site == "NVBR" ~ "North of Venoco Bridge",
    site == "NEC" ~ "South of Venoco Bridge",
    site == "NMC" ~ "Main Channel",
    site == "NPB" ~ "South of Phelps Bridge",
    site == "NPB1" ~ "North of Phelps Bridge",
    site == "NPB2" ~ "Phelps Road",
    site == "NWP" ~  "West Pond",
    site == "NDC" ~ "Devereux Creek"
  )) |> 
  # setting factor levels for site
  mutate(site_full = fct_reorder(site_full, med_sal, .fun = "median", .na_rm = TRUE),
         site_full = fct_rev(site_full))
```

## Class code

```{r}
#| label: visualizing-salinity
#| fig-width: 8
#| fig-height: 6

# base layer
salinity <- ggplot(data = aq_ins_clean,
       # x-axis: site
       # y-axis: mean salinity
       mapping = aes(x = site_full,
                     y = med_sal)) +
  # first layer: salinity measurements at each survey
  geom_jitter(height = 0,
              width = 0.1,
              shape = 21) +
  # wrapping x-axis text
  scale_x_discrete(labels = label_wrap(width = 10)) +
  # second layer: points to represent means
  stat_summary(fun = median,
               color = "red",
               size = 1) +
  # relabelling x- and y-axes
  labs(x = "Site",
       y = "Median salinity (ppt)",
       title = "Site salinity") +
  # different theme
  theme_bw()
```

```{r}
#| label: copepod-cleaning

# creating new object from clean aquatic inverts 
copepods <- aq_ins_clean |> 
  # filter to only include copepods
  filter(taxon == "copepod") |> 
  # log + 1 transform mean abundance per liter
  mutate(log_ave_lit = log(ave_lit + 1)) 
```

```{r}
#| label: visualizing-copepods
#| fig-width: 10
#| fig-height: 6

# base layer
copepods <- ggplot(data = copepods,
                   # x-axis: site
                   # y-axis: mean abundance/L
                   mapping = aes(x = site_full,
                                 y = log_ave_lit)) +
  # first layer: points to represent observations
  geom_jitter(height = 0,
              shape = 21,
              width = 0.1) +
  # second layer: points to represent medians
  stat_summary(geom = "point",
               fun = "median",
               size = 4,
               color = "blue") +
  # wrapping x-axis text
  scale_x_discrete(labels = label_wrap(10)) +
  # relabelling x- and y-axis, adding title
  labs(x = "Site",
       y = "log + 1 transformed average abundance/L",
       title = "Copepods") +
  # different theme
  theme_bw()
```


# 2. Problems

1. Visualizing differences across sites in major taxon abundance

a. Plan your figure

- create an object to store new plot
- x-axis should be `site_full`
- y-axis should be `log_avg_lit`
- geometry to represent average abundance/L: `geom_jitter()`
- geometry to represent medians: `stat_summary()`

b. Wrangle your data

```{r}
#| label: ostracod-cleaning

# creating new object from clean aquatic inverts
ostracods <- aq_ins_clean |> 
  # filtering to only include ostrapods
  filter(taxon == "ostracod") |> 
  # log + 1 transform mean abundance per liter
  mutate(log_ave_lit = log(ave_lit + 1))

slice_sample(ostracods, n = 10)
```

c. Code your figure

```{r}
#| label: visualizing-ostracod
#| fig-height: 6
#| fig-width: 8

# creating new object to store ostracod plot
ostracod_plot <- ggplot(data = ostracods,
                        # x-axis: full site name
                        # y-axis: log transformed average abundance/L
                        mapping = aes(x = site_full,
                                      y = log_ave_lit)) +
  # first layer: points to represent observations
  geom_jitter(height = 0,
              width = 0.1,
              shape = 21) +
  # second layer: points to represent median
  stat_summary(geom = "point",
               fun = "median",
               size = 4,
               color = "pink") +
  # wrapping x-axis labels to fit
  scale_x_discrete(labels = label_wrap(10)) +
  # relabeling axes and adding title
  labs(x = "Site",
       y = "log + 1 transformed average abundance/L",
       title = "Ostracods") +
  # custom theme
  theme_bw()

# displaying plot
ostracod_plot
```

d. Create a multipanel figure

```{r}
#| label: mulitpanel
#| fig-height: 10
#| fig-width: 10

# combining salinity, copepod, and ostracod plots into one figure
salinity / (copepods | ostracod_plot)
```

e. Interpret your figure

Copepod abundance is high at North of Venoco Bridge, South of Venoco Bridge, South of Phelps Bridge, and Devereux Creek. It is low at Main Channel, North of Phelps Bridge, and Phelps Road.

Ostracod abundance is high at South of Phelps Bridge and Devereux Creek. It is low at North of Venoco Bridge, South of Vencoco Bridge, Main Channel, North of Phelps Bridge, and Phelps Road.

The two highest levels of ostracod abundance are at South of Phelps Bridge and Devereux Creek. When looking at the median salinity for those two sites, they are both very low, Devereux Creek being the lowest. This may show that ostracods prefer low salinity.
